Teaching Low-Level Concurrency with \pebbles and \landslide

### 2016-08-26 sigcse-17/410.tex
sigcse: add some pretty pictures
sigcse: talk about heuristix
sigcse: shorten some things and lengthen others
sigcse: fill in a bunch of TODOs and anonymized citations

A second goal is encouraging students to
consider how ingredients are assembled into
abstractions.
%% Good rhetoric (below). -- bblum
There is a semantic gap
between the library-level \x{thr_create()}
call (analogous to \x{pthread_create()})
and the \x{thread_fork} system call provided
by the underlying kernel,
which takes no parameters,
makes no changes to the user's address space,
and cannot meaningfully
be invoked from C~code.
Because these abstractions are so different,
students must
understand each as an independent entity while
deciding how to bridge them.
A similar deliberate semantic gap exists between
\x{cond_wait()} and \x{deschedule()} (the system call
which suspends execution of a thread).

Student library code is linked against small
test programs provided by the course staff,
%producing one statically linked ELF executable
%per test program.}
%The test programs \shortversion{run on}{are bundled into a RAM-disk
%image and linked against}
which then run on a ``reference kernel'' provided in binary form.
%written by
%the course staff
%and provided in binary form.
The reference kernel's behavior is specified
% TODO: Anonymize this citation
in a twelve-page document~\cite{kspec}.
In addition to providing a reliable execution
substrate,
the reference kernel
schedules user-space threads created by
student code using a variety of
interleaving policies.
%\shortversion{}{In order to aid debugging of concurrency problems,
%the \x{misbehave()} system call enables
%students to manually select an interleaving policy.}

@@

The main execution and debugging
platform used in the class is
Wind River Simics~\cite{simics}.
%originally developed by the Swedish Institute of
%Computer Science and
%currently a product of Wind River.
Unlike some emulators,
which focus on fast execution of correct code,
Simics
provides very faithful bit-level support
not only for correct code %that behaves correctly
but also for kernels that accidentally ``abuse'' hardware.
%{but also for code that is ``incorrect within reason.''
%Occasionally student code manages to invoke an unsupported
%function of some hardware device or even to cause the
%simulator to crash, but this is quite rare, and in almost
%every case when the ``bad event'' is characterized it turns
%out that what the software requested from the hardware was
%a mistake.
%Meanwhile, inside its large envelope, Simics supports
%\textit{very} detailed models of various x86 processors.}
Unlike hardware virtualization environments,
Simics contains substantial debugger support:
single-stepping,
printing of source-level symbolic
expressions,
stack tracing,
display of TLB entries,
and even summaries of x86
%hardware-defined
descriptor tables.
%Unlike some symbolic debuggers,
%Simics is scriptable in Python,
%and it is possible to interpose user-written measurement
%frameworks to evaluate proposed hardware
%features~\shortversion{\cite{SIMFLEX}}{\cite{SIMFLEX,UW-GEMS,FeS2}}.
%\shortversion{}{Performance is very reasonable (Simics can simulate certain
%cases, such as a CPU halted awaiting a clock interrupt,
%much faster than real time).}
%\shortversion{}{A major advantage of using Simics over the QEMU emulator in particular
%is that QEMU only issues timer interrupts at basic-block boundaries,
%which dramatically undermines our goal of teaching students that threads can interleave with each other at any time.}

@@

\subsection{Grading}

@@

% TODO: Figure out how to left-justify these bullet poince.
\begin{itemize}
	\item Students learn more from bugs
		that are harder to find.
		In extreme cases, %code may do
		%almost everything right, but the last bug may
		subtle bugs may
		demonstrate a deep misunderstanding which requires a re-design.

@@

% TODO: Talk about P2 hurdle

A ``complete'' kernel can boot up and run the shell,
can pass 80\% of the basic system-call sanity test,
70\% of the solidity tests (which
invoke system calls in illegal
%or non-trivial
ways),
and two of the three long-running multi-feature tests.
%In addition,
%a complete kernel is certified by the authors to
Students also certify their thread libraries and kernels to
be free of ``concurrency shortcuts,''
to handle resource exhaustion,
and to avoid certain distasteful x86-specific code structures.

>>>>

A second goal is encouraging students to
consider how ingredients are assembled into
abstractions.
%% Good rhetoric (below). -- bblum
There is a semantic gap
between the library-level \x{thr_create()}
call (analogous to \x{pthread_create()})
and the \x{thread_fork} system call provided
by the underlying kernel,
which takes no parameters,
makes no changes to the user's address space,
and cannot meaningfully
be invoked from C~code.
Because these abstractions are so different,
students must
understand each as an independent entity while
deciding how to bridge them.
A similar deliberate semantic gap exists between
\x{cond_wait()} and \x{deschedule()}.
%(the system call which suspends execution of a thread).

Student library code is linked against small
test programs provided by the course staff,
%producing one statically linked ELF executable
%per test program.}
%The test programs \shortversion{run on}{are bundled into a RAM-disk
%image and linked against}
which then run on a ``reference kernel'' provided in binary form.
%written by
%the course staff
%and provided in binary form.
The reference kernel's behavior is specified
% TODO CAMREADY: deanonymize
in a twelve-page document~\cite{kspec-anonymized}.
In addition to providing a reliable execution
substrate,
the reference kernel
schedules user-space threads created by
student code using a variety of
interleaving policies.
%\shortversion{}{In order to aid debugging of concurrency problems,
%the \x{misbehave()} system call enables
%students to manually select an interleaving policy.}

@@

The main execution and debugging
platform used in the class is
Wind River Simics~\cite{simics}.
%originally developed by the Swedish Institute of
%Computer Science and
%currently a product of Wind River.
Unlike some emulators,
which focus on fast execution of correct code,
Simics
provides very faithful bit-level support
not only for correct code %that behaves correctly
but also for kernels that accidentally ``abuse'' hardware.
%{but also for code that is ``incorrect within reason.''
%Occasionally student code manages to invoke an unsupported
%function of some hardware device or even to cause the
%simulator to crash, but this is quite rare, and in almost
%every case when the ``bad event'' is characterized it turns
%out that what the software requested from the hardware was
%a mistake.
%Meanwhile, inside its large envelope, Simics supports
%\textit{very} detailed models of various x86 processors.}
Unlike hardware virtualization environments,
Simics contains substantial debugger support:
single-stepping,
printing of source-level symbolic
expressions,
stack tracing,
display of TLB entries,
and even summaries of x86
%hardware-defined
descriptor tables.
%Unlike some symbolic debuggers,
%Simics is scriptable in Python,
%and it is possible to interpose user-written measurement
%frameworks to evaluate proposed hardware
%features~\shortversion{\cite{SIMFLEX}}{\cite{SIMFLEX,UW-GEMS,FeS2}}.
%\shortversion{}{Performance is very reasonable (Simics can simulate certain
%cases, such as a CPU halted awaiting a clock interrupt,
%much faster than real time).}
%\shortversion{}{A major advantage of using Simics over the QEMU emulator in particular
%is that QEMU only issues timer interrupts at basic-block boundaries,
%which dramatically undermines our goal of teaching students that threads can interleave with each other at any time.}
It also provides instruction and memory access tracing and checkpointing/rewinding, both of which Landslide employs.

@@

\subsection{Grading}
\label{sec:grading}

@@

\begin{itemize}
	\item Students learn more from bugs
		that are harder to find.
		In extreme cases, %code may do
		%almost everything right, but the last bug may
		subtle bugs may
		demonstrate a deep misunderstanding which requires a re-design.

@@

The thread library hurdle requires students to pass a suite of five test cases,
which test mutex locking/unlocking and thread creation, exit, and joining in moderately stressful ways.
%
A hurdle-passing kernel can boot up and run the shell,
can pass 80\% of the basic system-call sanity test,
70\% of the solidity tests (which
invoke system calls in illegal
%or non-trivial
ways),
and two of the three long-running multi-feature tests.
%In addition,
%a complete kernel is certified by the authors to
Students also certify their thread libraries and kernels to
be free of ``concurrency shortcuts,''
to handle resource exhaustion,
and to avoid certain distasteful x86-specific code structures.

### 2016-08-26 sigcse-17/landslide.tex
sigcse: fill in a bunch of TODOs and anonymized citations

>>>>

Apart from which test program to run, \landslide's user interface requires just one option:
students specify how much time they are willing to wait.
%Considering this CPU budget,
\landslide then uses state space estimation \cite{estimation} to predict
% TODO CAMREADY: deanonymize
which combinations of preemption points will produce appropriately-sized state spaces for the given CPU-time budget \cite{quicksand-anonymized}.
If no bug is found before time runs out, \landslide displays a progress report summarizing the estimated remaining time for each state space,
so ambitious students can decide whether to increase the time limit and retry.
